---
title: "explore_obis"
format: html
editor_options: 
  chunk_output_type: console
---

## Goal

Explore OBIS occurrence data for ranking species by number of occurrences, spatial/temporal distribution and taxonomic importance (e.g., trophic level, extinction risk). This will facilitate a short list of species for applying initial species distribution models (SDMs).

Source: * [Data access | Ocean Biodiversity Information System](https://obis.org/data/access/#)

## Read OBIS parquet

* [BigelowLab/fullobis](https://github.com/BigelowLab/fullobis)
    Simple Access to the Full OBIS Dataset with R
* [Diversity indicators using OBIS data](https://iobis.github.io/notebook-diversity-indicators/)

OBIS

* [8 Data Visualization - The OBIS manual](https://manual.obis.org/dataviz.html#obisindicators-calculating-visualizing-spatial-biodiversity-using-data-from-obis)
* [api.obis.org](https://api.obis.org/)
* [External data sources | OBIS](https://obis.org/data/externalsources/)

```{r}
# libraries ----
librarian::shelf(
  arrow, dplyr, 
  BigelowLab/fullobis,
  glue, here, mapview, readr, sf)

# variables -----
dir_data     <- "/Users/bbest/My Drive/projects/mbon-sdm/data"
obis_prq     <- glue("{dir_data}/raw/obis.org/obis_20230726.parquet")      #  4.9 GB
obis_csv     <- glue("{dir_data}/raw/obis.org/obis_20230726_subcols.csv")  # 16.6 GB
obis_sum_csv <- here("data/obis_summary.csv")                              #  4.1 MB
obis_top_csv <- here("data/obis_top.csv")                                  #      MB

# obis_prq <- "/Users/bbest/Downloads/obis_20230726.parquet"

# fullobis setup
set_root(glue("{dir_data}/raw/obis.org"))
set_useragent("http://github.com/bbest")

# checks ----
stopifnot(file.exists(obis_prq))
```


```{r}
#| eval: False

# fetch OBIS parquet
# ok       <- fetch_obis()
# obis_prq <- get_current()

# open and read all or parts of the file
x <- open_dataset(obis_prq)
x
# FileSystemDataset with 1 Parquet file
# id: string
# dataset_id: string
# decimalLongitude: float
# decimalLatitude: float
# date_start: int64
# date_mid: int64
# date_end: int64
# date_year: int64
# scientificName: string
# originalScientificName: string
# minimumDepthInMeters: float
# maximumDepthInMeters: float
# depth: float
# coordinateUncertaintyInMeters: float
# flags: string
# dropped: bool
# absence: bool
# shoredistance: string
# bathymetry: string
# sst: string
# sss: string
# marine: bool
# brackish: bool
# freshwater: bool
# terrestrial: bool
# taxonRank: string
# AphiaID: int64
# redlist_category: string
# superdomain: string
# domain: string
# kingdom: string
# subkingdom: string
# infrakingdom: string
# phylum: string
# phylum_division: string
# subphylum_subdivision: string
# subphylum: string
# infraphylum: string
# parvphylum: string
# gigaclass: string
# megaclass: string
# superclass: string
# class: string
# subclass: string
# infraclass: string
# subterclass: string
# superorder: string
# order: string
# suborder: string
# infraorder: string
# parvorder: string
# superfamily: string
# family: string
# subfamily: string
# supertribe: string
# tribe: string
# subtribe: string
# genus: string
# subgenus: string
# section: string
# subsection: string
# series: string
# species: string
# subspecies: string
# natio: string
# variety: string
# subvariety: string
# forma: string
# subforma: string
# superdomainid: int64
# domainid: int64
# kingdomid: int64
# subkingdomid: int64
# infrakingdomid: int64
# phylumid: int64
# phylum_divisionid: int64
# subphylum_subdivisionid: int64
# subphylumid: int64
# infraphylumid: int64
# parvphylumid: int64
# gigaclassid: int64
# megaclassid: int64
# superclassid: int64
# classid: int64
# subclassid: int64
# infraclassid: int64
# subterclassid: int64
# superorderid: int64
# orderid: int64
# suborderid: int64
# infraorderid: int64
# parvorderid: int64
# superfamilyid: int64
# familyid: int64
# subfamilyid: int64
# supertribeid: int64
# tribeid: int64
# subtribeid: int64
# genusid: int64
# subgenusid: int64
# sectionid: int64
# subsectionid: int64
# seriesid: int64
# speciesid: int64
# subspeciesid: int64
# natioid: int64
# varietyid: int64
# subvarietyid: int64
# formaid: int64
# subformaid: int64
# type: string
# modified: string
# language: string
# license: string
# rightsHolder: string
# accessRights: string
# bibliographicCitation: string
# references: string
# institutionID: string
# collectionID: string
# datasetID: string
# institutionCode: string
# collectionCode: string
# datasetName: string
# ownerInstitutionCode: string
# basisOfRecord: string
# informationWithheld: string
# dataGeneralizations: string
# dynamicProperties: string
# materialSampleID: string
# occurrenceID: string
# catalogNumber: string
# occurrenceRemarks: string
# recordNumber: string
# recordedBy: string
# recordedByID: string
# individualCount: string
# organismQuantity: string
# organismQuantityType: string
# sex: string
# lifeStage: string
# reproductiveCondition: string
# behavior: string
# establishmentMeans: string
# occurrenceStatus: string
# preparations: string
# disposition: string
# otherCatalogNumbers: string
# associatedMedia: string
# associatedReferences: string
# associatedSequences: string
# associatedTaxa: string
# organismID: string
# organismName: string
# organismScope: string
# associatedOccurrences: string
# associatedOrganisms: string
# previousIdentifications: string
# organismRemarks: string
# eventID: string
# parentEventID: string
# samplingProtocol: string
# sampleSizeValue: string
# sampleSizeUnit: string
# samplingEffort: string
# eventDate: string
# eventTime: string
# startDayOfYear: string
# endDayOfYear: string
# year: string
# month: string
# day: string
# verbatimEventDate: string
# habitat: string
# fieldNumber: string
# fieldNotes: string
# eventRemarks: string
# locationID: string
# higherGeographyID: string
# higherGeography: string
# continent: string
# waterBody: string
# islandGroup: string
# island: string
# country: string
# countryCode: string
# stateProvince: string
# county: string
# municipality: string
# locality: string
# verbatimLocality: string
# verbatimElevation: string
# minimumElevationInMeters: string
# maximumElevationInMeters: string
# verbatimDepth: string
# minimumDistanceAboveSurfaceInMeters: string
# maximumDistanceAboveSurfaceInMeters: string
# locationAccordingTo: string
# locationRemarks: string
# verbatimCoordinates: string
# verbatimLatitude: string
# verbatimLongitude: string
# verbatimCoordinateSystem: string
# verbatimSRS: string
# geodeticDatum: string
# coordinatePrecision: string
# pointRadiusSpatialFit: string
# footprintWKT: string
# footprintSRS: string
# footprintSpatialFit: string
# georeferencedBy: string
# georeferencedDate: string
# georeferenceProtocol: string
# georeferenceSources: string
# georeferenceVerificationStatus: string
# georeferenceRemarks: string
# geologicalContextID: string
# earliestEonOrLowestEonothem: string
# latestEonOrHighestEonothem: string
# earliestEraOrLowestErathem: string
# latestEraOrHighestErathem: string
# earliestPeriodOrLowestSystem: string
# latestPeriodOrHighestSystem: string
# earliestEpochOrLowestSeries: string
# latestEpochOrHighestSeries: string
# earliestAgeOrLowestStage: string
# latestAgeOrHighestStage: string
# lowestBiostratigraphicZone: string
# highestBiostratigraphicZone: string
# lithostratigraphicTerms: string
# group: string
# formation: string
# member: string
# bed: string
# identificationID: string
# identifiedBy: string
# identifiedByID: string
# dateIdentified: string
# identificationReferences: string
# identificationRemarks: string
# identificationQualifier: string
# identificationVerificationStatus: string
# typeStatus: string
# taxonID: string
# scientificNameID: string
# acceptedNameUsageID: string
# parentNameUsageID: string
# originalNameUsageID: string
# nameAccordingToID: string
# namePublishedInID: string
# taxonConceptID: string
# acceptedNameUsage: string
# parentNameUsage: string
# originalNameUsage: string
# nameAccordingTo: string
# namePublishedIn: string
# namePublishedInYear: string
# higherClassification: string
# specificEpithet: string
# infraspecificEpithet: string
# verbatimTaxonRank: string
# scientificNameAuthorship: string
# vernacularName: string
# nomenclaturalCode: string
# taxonomicStatus: string
# nomenclaturalStatus: string
# taxonRemarks: string

# columns to keep
col_keep = c(
  "id",
  "phylum",
  "class",
  "scientificName",
  "AphiaID",
  "date_year",
  "date_mid",
  "eventDate",
  "decimalLongitude",
  "decimalLatitude",
  "depth",
  "individualCount",
  "flags",
  "dropped",
  "absence")
system.time({
  y <- x |> 
    select(all_of(col_keep)) |> 
    collect() |> 
    as_tibble() |> 
    filter(!is.na(individualCount), !is.na(date_year))
})
# before extra columns (class, AphiaID, date_mid, eventDate, depth, flags, dropped, absence):
#      user   system  elapsed 
#   246.842 1052.350 1777.397   # /60    = 29.6 min
# after:
#      user    system   elapsed 
#   719.556  7455.860 13976.147 # /60/60 = 3.9  hrs

system.time({
  write_csv(y, obis_csv)
})
#   user  system elapsed                                                                                                                                             
# 37.480  14.787  16.088

y
# before extra columns (class, AphiaID, date_mid, eventDate, depth, flags, dropped, absence):
#   A tibble: 27,101,348 × 7
#      id                decimalLongitude decimalLatitude scientificName phylum date_year
#      <chr>                        <dbl>           <dbl> <chr>          <chr>      <int>
#    1 ab2b3d73-0437-42…           -177.            0.811 Pycnochromis … Chord…      2006
#    2 d0b85b79-3763-47…           -177.            0.811 Pycnochromis … Chord…      2006
#    3 1705ac8c-4903-4b…           -177.            0.811 Pycnochromis … Chord…      2006
#    4 7960289e-ddaf-47…            132.            2.92  Gymnothorax f… Chord…      2008
#    5 8e621c79-a0fd-4a…             10.5          56.9   Dipolydora co… Annel…      2004
#    6 f959831d-2189-49…             10.5          56.9   Dipolydora co… Annel…      2004
#    7 b28e4b42-74fe-4a…           -177.            0.811 Pycnochromis … Chord…      2006
#    8 008d1ba6-a337-4d…           -177.            0.811 Pycnochromis … Chord…      2006
#    9 3e37c8c0-357a-49…           -177.            0.811 Pycnochromis … Chord…      2006
#   10 28f7e5e6-9fe8-4e…           -177.            0.790 Monotaxis gra… Chord…      2006
#   ℹ 27,101,338 more rows
#   ℹ 1 more variable: individualCount <chr>
# after:
#   A tibble: 27,101,348 × 15
#      id         phylum class scientificName AphiaID date_year date_mid eventDate decimalLongitude decimalLatitude depth individualCount flags
#      <chr>      <chr>  <chr> <chr>            <int>     <int>  <int64> <chr>                <dbl>           <dbl> <dbl> <chr>           <chr>
#    1 ab2b3d73-… Chord… Tele… Pycnochromis … 1595014      2006     1e12 2006-01-…           -177.            0.811  14   7               {}   
#    2 d0b85b79-… Chord… Tele… Pycnochromis … 1595014      2006     1e12 2006-01-…           -177.            0.811  14   4               {}   
#    3 1705ac8c-… Chord… Tele… Pycnochromis … 1595014      2006     1e12 2006-01-…           -177.            0.811  14   5               {}   
#    4 7960289e-… Chord… Tele… Gymnothorax f…  217499      2008     1e12 2008-09-…            132.            2.92   NA   1               {NO_…
#    5 8e621c79-… Annel… Poly… Dipolydora co…  131117      2004     1e12 2004-04-…             10.5          56.9    11.6 1.0             {}   
#    6 f959831d-… Annel… Poly… Dipolydora co…  131117      2004     1e12 2004-04-…             10.5          56.9    11.9 1.0             {}   
#    7 b28e4b42-… Chord… Tele… Pycnochromis … 1595014      2006     1e12 2006-01-…           -177.            0.811  14   20              {}   
#    8 008d1ba6-… Chord… Tele… Pycnochromis … 1595014      2006     1e12 2006-01-…           -177.            0.811  14   34              {}   
#    9 3e37c8c0-… Chord… Tele… Pycnochromis … 1595014      2006     1e12 2006-01-…           -177.            0.811  14   20              {}   
#   10 28f7e5e6-… Chord… Tele… Monotaxis gra…  218584      2006     1e12 2006-01-…           -177.            0.790  14   1               {}   
#   ℹ 27,101,338 more rows
#   ℹ 2 more variables: dropped <lgl>, absence <lgl>

system.time({
  y_sum <- y |> 
    # TODO: get Kingdom
    group_by(phylum, class, scientificName, AphiaID) |>
    summarize(
      year_min = min(date_year),
      n_years  = n_distinct(date_year),
      year_max = max(date_year),
      n_obs    = n(),
      n_indiv  = sum(as.numeric(individualCount), na.rm=T),
      .groups = "drop") |> 
  arrange(desc(n_obs), phylum, class, scientificName, AphiaID)
})
#    user  system elapsed 
#   3.792   0.960   5.366
  
table(is.na(y_sum$n_indiv))
# before na.rm=T:
#    FALSE  TRUE 
#   88,999    74
# after na.rm=T:
#    FALSE 
#   89,073
#   89,216
y_sum
# A tibble: 89,073 × 7
#    phylum     scientificName           year_min n_years year_max  n_obs     n_indiv
#    <chr>      <chr>                       <int>   <int>    <int>  <int>       <dbl>
#  1 Chordata   Fulmarus glacialis           1938      56     2021 892182    3851813 
#  2 Chordata   Uria aalge                   1966      54     2021 625780    1603709 
#  3 Chordata   Gadus morhua                 1782     102     2020 594939   35768843.
#  4 Chordata   Merlangius merlangus         1921      65     2021 493998  740174517.
#  5 Chordata   Melanogrammus aeglefinus     1913      72     2013 466065  678710462.
#  6 Chordata   Morus bassanus               1965      56     2021 461615     954192 
#  7 Chordata   Rissa tridactyla             1965      56     2021 428197    1885176 
#  8 Chordata   Clupea harengus              1757     154     2019 396647 1938019751.
#  9 Arthropoda Homarus americanus           1874      65     2014 384192    3485011.
# 10 Chordata   Limanda limanda              1927      73     2021 345898  251012411.
# ℹ 89,063 more rows

n_distinct(y_sum$phylum)
# 61

n_distinct(y_sum$class)
# 174

write_csv(y_sum, obis_sum_csv)
```


```{r}
d_o <- read_csv(obis_sum_csv)
dim(d_o)
# 89216     9

d_o_top <- d_o |> 
  arrange(desc(n_obs), phylum, class) |>
  group_by(phylum, class) |>
  summarize(
    scientificName = first(scientificName), 
    AphiaID        = first(AphiaID), 
    year_min       = first(year_min), 
    n_years        = first(n_years), 
    year_max       = first(year_max), 
    n_obs          = first(n_obs), 
    n_indiv        = first(n_indiv), 
    .groups = "drop")
dim(d_o_top)
# 233   9
write_csv(d_o_top, obis_top_csv)

DT::datatable(d_o_top)
```



* [Fulmarus glacialis | WoRMS](https://marinespecies.org/aphia.php?p=taxdetails&id=137195)
  - Class: Aves
  - Vernaculars: northern fulmar, fulmar
* [Melanogrammus aeglefinus | WoRMS](https://marinespecies.org/aphia.php?p=taxdetails&id=126437)
  - Class: Actinopterygii
  - Vernaculars: haddock
* [Homarus americanus | WoRMS](https://marinespecies.org/aphia.php?p=taxdetails&id=156134)
  - Class: Malacostraca
  - Vernaculars: northern lobster, maine lobster, lobster, American lobster, American clawed lobster
    
